Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-28 overleaf/model.tex
Finished edits to 3.

This section presents a model to reason about the cost of execution of MPC programs, including accounting for amortization.
We define the assumptions and setting in~\secref{sec:mpc}. We proceed to define the scheduling problem in~\secref{sec:problem}, 
which we expected to be able to solve optimally. \secref{sec:np} show that the problem is NP-hard via a reduction to the 
Shortest Common Supersequence (SCS) problem. Despite the negative general result, we expect the formulation in terms of SCS to 
be useful as sequences are short and few in practice.

@@

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion.
%\ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item There are two tiers of MPC instructions, local and remote. A local instruction (essentially just ADD) has cost $\beta$
and a remote instruction (e.g., MUX, MUL, SHL, etc.) has cost $\alpha$, where $\alpha >> \beta$. We assume that all remote
instructions have the same cost.
%\ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item  In MPC frameworks, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
Following Amdahl's law~\ana{citation!}, we write $\alpha = \frac{1}{s}p\alpha + (1-p)\alpha$, where $p$ is the fraction of execution time that benefits from amortization and $(1-p)$
is the fraction that does not, and $s$ is the available resource. Thus, $n\alpha = \frac{n}{s}p\alpha + n(1-p)\alpha$.
For the purpose of the model we assume that $s$ is large enough and the term $\frac{n}{s}p\alpha$ amounts to a \emph{fixed cost} incurred regardless of
whether $n$ is $10,000$ or just $1$. (This models the cost of preparing and sending a packet from party A to party B for example.) The bottom line is that amortized execution
of $n$ operations is $f(n) = \alpha_\mathit{fix} + n\alpha_{var}$ in contrast to unamortized execution $g(n) = n\alpha_\mathit{fix} + n\alpha_{var}$ and, of course,
$\alpha_\mathit{fix} << n\alpha_\mathit{fix}$.
%We assume infinite parallel capacity---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely $\alpha$ or $\beta$.
%This is a standard assumption in Cryptographic Parallel RAM. ABY presents empirical support for this assumption~\ana{Add citations. PRAM, ABY.}
%\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions can be amortized in a single SIMD instruction that costs $\alpha_\mathit{fix} + 2\alpha_\mathit{var}$, however a MUL and a MUX instruction
still cost $2\alpha_\mathit{fix} + 2\alpha_\mathit{var}$ even when scheduled ``in parallel''.
%MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
%2 MUL instructions scheduled in parallel benefit from amortization and cost $\alpha$, however a MUL and a MUX instructions scheduled
%in parallel still cost $2\alpha$.
%\ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

>>>>

This section presents a model to reason about the cost of execution of MPC programs, including accounting for amortization.
We define the assumptions and setting in~\secref{sec:mpc}. We proceed to define the scheduling problem in~\secref{sec:problem}, 
which we expected to be able to solve optimally. \secref{sec:np} shows that the problem is NP-hard via a reduction to the 
Shortest Common Supersequence (SCS) problem. Despite the negative general result, we expect the formulation in terms of SCS to 
be useful as sequences are short and few in practice.

@@

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion.
%\ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item There are two tiers of MPC instructions, local and remote. A local instruction (essentially just ADD) has cost $\beta$
and a remote instruction (e.g., MUX, MUL, SHL, etc.) has cost $\alpha$, where $\alpha >> \beta$. We assume that all remote
instructions have the same cost.
%\ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item  In MPC frameworks, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
Following Amdahl's law~\ana{citation!}, we write $\alpha = \frac{1}{s}p\alpha + (1-p)\alpha$, where $p$ is the fraction of execution time that benefits from amortization and $(1-p)$
is the fraction that does not, and $s$ is the available resource. Thus, $n\alpha = \frac{n}{s}p\alpha + n(1-p)\alpha$.
For the purpose of the model we assume that $s$ is large enough and the term $\frac{n}{s}p\alpha$ amounts to a \emph{fixed cost} incurred regardless of
whether $n$ is $10,000$ or just $1$. (This models the cost of preparing and sending a packet from party A to party B for example.) Therefore, amortized execution
of $n$ operations is $f(n) = \alpha_\mathit{fix} + n\alpha_{var}$ in contrast to unamortized execution $g(n) = n\alpha_\mathit{fix} + n\alpha_{var}$ and, of course.
Clearly $\alpha_\mathit{fix} << n\alpha_\mathit{fix}$ and since fixed cost dominates variable cost (particularly for remote operations), we have $f(n) << g(n)$.
%We assume infinite parallel capacity---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely $\alpha$ or $\beta$.
%This is a standard assumption in Cryptographic Parallel RAM. ABY presents empirical support for this assumption~\ana{Add citations. PRAM, ABY.}
%\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions can be amortized in a single SIMD instruction that costs $\alpha_\mathit{fix} + 2\alpha_\mathit{var}$, however a MUL and a MUX instruction
still cost $2\alpha_\mathit{fix} + 2\alpha_\mathit{var}$ even when scheduled ``in parallel''.\footnote{This is not strictly true, but assuming it, e.g. as in 
\cite{Ishaq2019, Demmler2015ABYA, Mohassel2018}, helps simplify the problem.}
%MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
%2 MUL instructions scheduled in parallel benefit from amortization and cost $\alpha$, however a MUL and a MUX instructions scheduled
%in parallel still cost $2\alpha$.
%\ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

### 2022-04-28 overleaf/prob_stmt_np_hardness.tex
Finished edits to 3.

\begin{enumerate}
    \item All $P$'s consist of MPC instructions of the same kind, e.g., all MUL, or ADD, etc. 
    \item Def-use dependencies of the graph $G(V, E)$ should be preserved i.e. if instructions $S_i, S_j, i < j$ form a def-use i.e. an edge exists from $S_i$ to $S_j$ in $G$, then they can only be mapped to $P_{i'}, P_{j'}$ such that $i' < j'$.
\end{enumerate}

@@

In order to benefit from parallelization/amortization, we must schedule two or more $A$-instructions in the same parallel node (or two or more $M$-instructions in the same parallel node). Recall that we also assume that scheduling $A$-instructions in parallel with $M$-instruction does not benefit from amortization\footnote{this is not strictly true, but assuming it, e.g. as in \cite{Ishaq2019, Demmler2015ABYA, Mohassel2018}, helps simplify the problem.}. It incurs the exact same cost as scheduling the $A$-instructions in a node $P_A$, scheduling the $M$-instructions in a node $P_M$, and having $P_A$ precede $P_M$ in the parallel schedule. We use the following cost model:

@@

The cost of schedule $S$ is
\begin{equation}
    \mathit{cost}(S) = \sum_{i=1}^n \mathit{cost}(S_i)
\end{equation}

where $cost(S_i) = \alpha$ or $\beta$. %if $S_i$ is an $A$-instruction and $cost(S_i) = \beta$ if $S_i$ is an $M$-instruction. 
Similarly, the cost of schedule $P$ is

@@

Each $P_i$ may contain multiple instructions, and $\mathit{cost}(P_i)$ is amortized.
Thus, according to our model $\mathit{cost}(P_i) = \alpha_\mathit{fix} + |P_i|\alpha_\mathit{var}$ if $P_i$ stores $\alpha$-instructions, 
or $\mathit{cost}(P_i) = \beta_\mathit{fix} + |P_i|\beta_\mathit{var}$ if it stores $\beta$-instructions.
%if $P_i$ consists of $A$-instructions only, it is $\beta$ if $P_i$ consists of $M$-instructions only, and it is $(\alpha + \beta)$ if $P_i$ mixes $A$-instructions and $M$-instructions. 
Our goal is to construct a parallel schedule $P$ that reduces the program cost (when compared to cost of $S$), possibly an optimal schedule. 
Originally we hoped that the problem is simper and computation of the optimal schedule is tractable. Unfortunately, the optimal schedule turns out to be NP-hard via a reduction 
to the Shortest Common Supersequence problem.

@@

Note that we consider a linearized MPC schedule $S$ above for ease of exposition only. In our tool-chain we use an MPC Source control flow graph (CFG) $G'(V', E')$ along with def-use graph $G(V,E)$ to construct $P$. %The argument becomes slightly more involved when dealing with a graph $G'$ that may contain cycles.

@@

To prove that optimal scheduling is an NP-Hard problem, we consider the following convenient representation. An MPC program is represented as a set of sequences $S = \{S_1, \dots, S_n\}$. Each element $S_i \in S$ is a tuple. The items of the tuple $S_i$ are operations, i.e. $A$ or $M$ instructions, that have to be executed in order (operations depend on previous operations i.e. $S_i[j], j > 1$ depends $S_i[j-1]$). However, the sequences $S_i, 1 \leq i \leq n$ themselves, can overlap each other in any way i.e. distinct sequences can be executed in parallel. We argue that an MPC program can be transformed into such collection of sequences by traversing the circuit for each pair of input and output values. \ishaq{I am not sure how this will be done, see \cref{fig:sequence_construction}}

As an example, consider the MPC program consisting of the following three sequences, all made up of two distinct $\alpha$-instructions $M_1$ and $M_2$. The right arrow indicates a def-use \emph{dependence}, meaning that the source node must execute before the target node:

@@

The problem is to find a schedule $P$ with \emph{minimal cost}. For example, a schedule with minimal cost for the sequences above is

$$ 
M_1(1), M_1(2) \; ;\; M_1(2) \; ; \; M_2(1), M_2(3) \; ; \; M_1(1), M_1(2), M_1(3) \; ; \;  M_2(3)
$$

The parentheses above indicate the sequence where the instruction comes from: (1), (2), or (3). 
Cost of schedule $P$ is computed using \cref{eq_par_schedule_cost} above and it amounts to $5\alpha_\mathit{fix} + 9\alpha_\mathit{var}$.

The problem of finding a schedule $P$ with a minimal $cost(P)$ is shown to be NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, a known NP-Hard problem\cite{Maier1978, Vazirani2010}. The shortest common supersequence problem is as follows: {\it given two or more sequences find the the shortest sequence that contains all of the original sequences.} This can be solved in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the number of sequences. For our problem $n$ is the maximum length of a sequence and $k$ is the number of total number of sequences. We can immediately see that the optimal schedule is the shortest schedule, since the shortest schedule minimizes the fixed cost while the variable cost remains the same.

To formalize the reduction, suppose $P$ is a schedule with minimal cost (computed by a black-box algorithm). 
%We can derive a schedule $P'$ with the same cost as $P$, by mapping each mixed node $P_i \in P$ to two consecutive nodes in $P'$: an $A$-instruction node followed by an $M$-instruction node. 
Clearly, $P$, which now is a sequence of $M_1$ and $M_2$ nodes, is a supersequence of each sequence $S_i$, i.e., $P$ is a common supersequence of $S_1 \dots S_n$. It is also a shortest common supersequence. 
%To see this, let $X$ and $Y$ denote, respectively, the number of $A$ and $M$ nodes in $P'$. 
The cost of $\mathit{cost}(P) = L\alpha_\mathit{fix}+N\alpha_\mathit{var}$ where $L$ is the length of $P$ and $N$ is the total number of instructions across all sequences. %$X \cdot \alpha + Y \cdot \beta$. 
Now suppose, there exist a shorter common supersequence $P'$ of length $L'$. $\mathit{cost}(P') < \mathit{cost}(P)$ since $L'\alpha_\mathit{var} + N\alpha_\mathit{var} < L\alpha_\mathit{var} + N\alpha_\mathit{var}$, however, this is a contradiction to the assumption that $P$ has the lowest cost. \qed
%that consists of $X'$ nodes of type $A$-instructions $Y'$ nodes of type $M$-instructions. Since $P''$ is shorter than $P'$, therefore $X' + Y' < X + Y$, and $X' \cdot \alpha + Y' \cdot \beta < X \cdot \alpha + Y \cdot \beta$ i.e. $\mathit{cost}(P'') < \mathit{cost}(P')$. But $\mathit{cost}(P') = \mathit{cost}(P)$ and $\mathit{cost}(P)$ is the optimal cost.  Therefore $\mathit{cost}(P'') < \mathit{cost}(P')$ is contradiction and no such $P''$ exists. \qed

>>>>

\begin{enumerate}
    \item All $P_i$'s consist of MPC instructions of the same kind, e.g., all MUL, MUX, ADD, etc. 
    \item Def-use dependencies of the graph $G(V, E)$ should be preserved i.e. if instructions $S_i, S_j, i < j$ form a def-use i.e. an edge exists from $S_i$ to $S_j$ in $G$, then they can only be mapped to $P_{i'}, P_{j'}$ such that $i' < j'$.
\end{enumerate}

@@

%In order to benefit from parallelization/amortization, we must schedule two or more $A$-instructions in the same parallel node $P_i$ (or two or more $M$-instructions in the same parallel node). Recall that we also assume that scheduling $A$-instructions in parallel with $M$-instruction does not benefit from amortization\footnote{this is not strictly true, but assuming it, e.g. as in \cite{Ishaq2019, Demmler2015ABYA, Mohassel2018}, helps simplify the problem.}. It incurs the exact same cost as scheduling the $A$-instructions in a node $P_A$, scheduling the $M$-instructions in a node $P_M$, and having $P_A$ precede $P_M$ in the parallel schedule. We use the following cost model:

@@

The cost of schedule $S$ is
\begin{equation}
    \mathit{cost}(S) = \sum_{i=1}^n \mathit{cost}(S_i) = L_\alpha \alpha_\mathit{fix} + L_\beta \beta_\mathit{fix} + L_\alpha \alpha_\mathit{var} + L_\beta \beta_\mathit{var}
\end{equation}

where $L_\alpha$ is the number of $\alpha$-instructions and $L_\beta$ is the number of $\beta$ ones.
(We used this formula to compute the cost of the unrolled MPC Source program in~\secref{sec:overview}.)
%$cost(S_i) = \alpha$ or $\beta$. if $S_i$ is an $A$-instruction and $cost(S_i) = \beta$ if $S_i$ is an $M$-instruction. 
The cost of schedule $P$ is more interesting:

@@

Each $P_i$ may contain multiple instructions, and $\mathit{cost}(P_i)$ is amortized.
Thus, according to our model $\mathit{cost}(P_i) = \alpha_\mathit{fix} + |P_i|\alpha_\mathit{var}$ if $P_i$ stores $|P_i|$ $\alpha$-instructions, 
or $\mathit{cost}(P_i) = \beta_\mathit{fix} + |P_i|\beta_\mathit{var}$ if it stores $\beta$-instructions. 
(Similarly, we used this formula to compute the cost of the Optimized MPC Source program in~\secref{sec:overview}.)
%if $P_i$ consists of $A$-instructions only, it is $\beta$ if $P_i$ consists of $M$-instructions only, and it is $(\alpha + \beta)$ if $P_i$ mixes $A$-instructions and $M$-instructions.

Our goal is to construct a parallel schedule $P$ that reduces the program cost (when compared to cost of $S$), possibly an optimal schedule. 
Originally we hoped that the problem is simper and computation of the optimal schedule is tractable. Unfortunately, the optimal schedule turns out to be NP-hard via a reduction 
to the Shortest Common Supersequence problem.

@@

%Note that we consider a linearized MPC schedule $S$ above for ease of exposition only. In our tool-chain we use an MPC Source control flow graph (CFG) $G'(V', E')$ along with def-use graph $G(V,E)$ to construct $P$. %The argument becomes slightly more involved when dealing with a graph $G'$ that may contain cycles.

@@

To prove that optimal scheduling is an NP-Hard problem, we consider the following convenient representation. An MPC program is represented as a set of sequences $\{s_1, \dots, s_n\}$ of operations. 
%Each element $S_i \in S$ is a tuple. The items of the tuple $S_i$ are operations, i.e. $A$ or $M$ instructions, that have to be executed in order (
In each sequence $s_i$ operations depend on previous operations via a def-use i.e. $s_i[j], j > 1$ depends on $s_i[j-1]$. 
%However, the sequences $S_i, 1 \leq i \leq n$ themselves, can overlap each other in any way i.e. distinct sequences can be executed in parallel. We argue that an MPC program can be transformed into such collection of sequences by traversing the circuit for each pair of input and output values. \ishaq{I am not sure how this will be done, see \cref{fig:sequence_construction}}

As an example, consider the MPC program consisting of the following three sequences, all made up of two distinct $\alpha$-instructions $M_1$ and $M_2$, e.g., a $M_1$ is MUL and $M_2$ is MUX. The right arrow indicates a def-use \emph{dependence}, meaning that the source node must execute before the target node:

@@

The problem is to find a schedule $P$ with \emph{minimal cost}. For example, a schedule with minimal cost for the sequences above is 
\[ 
M_1(1), M_1(2) \; ;\; M_1(2) \; ; \; M_2(1), M_2(3) \; ; \; M_1(1), M_1(2), M_1(3) \; ; \;  M_2(3)
\]
The parentheses above indicate the sequence where the instruction comes from: (1), (2), or (3). 
Cost of schedule $P$ is computed using \cref{eq_par_schedule_cost} above and it amounts to $5\alpha_\mathit{fix} + 9\alpha_\mathit{var}$.

The problem of finding a schedule $P$ with a minimal $cost(P)$ is shown to be NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, a known NP-Hard problem\cite{Maier1978, Vazirani2010}. The shortest common supersequence problem is as follows: {\it given two or more sequences find the the shortest sequence that contains all of the original sequences.} This can be solved in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the number of sequences. For our problem $n$ is the maximum length of a sequence and $k$ is the number of instructions of sequences. We can immediately see that the optimal schedule is the shortest schedule, since the shortest schedule minimizes the fixed cost while the variable cost remains the same.

To formalize the reduction, suppose $P$ is a schedule with minimal cost (computed by a black-box algorithm). 
%We can derive a schedule $P'$ with the same cost as $P$, by mapping each mixed node $P_i \in P$ to two consecutive nodes in $P'$: an $A$-instruction node followed by an $M$-instruction node. 
Clearly $P$ is a supersequence of each sequence $s_i$, i.e., $P$ is a common supersequence of $s_1 \dots s_n$. It is also a shortest common supersequence. 
%To see this, let $X$ and $Y$ denote, respectively, the number of $A$ and $M$ nodes in $P'$. 
The cost of $\mathit{cost}(P) = L\alpha_\mathit{fix}+N\alpha_\mathit{var}$ where $L$ is the length of $P$ and $N$ is the total number of instructions across all sequences. %$X \cdot \alpha + Y \cdot \beta$. 
Now suppose, there exist a shorter common supersequence $P'$ of length $L'$. $\mathit{cost}(P') < \mathit{cost}(P)$ since $L'\alpha_\mathit{var} + N\alpha_\mathit{var} < L\alpha_\mathit{var} + N\alpha_\mathit{var}$, contradicting the assumption that $P$ has the lowest cost. \qed
%that consists of $X'$ nodes of type $A$-instructions $Y'$ nodes of type $M$-instructions. Since $P''$ is shorter than $P'$, therefore $X' + Y' < X + Y$, and $X' \cdot \alpha + Y' \cdot \beta < X \cdot \alpha + Y \cdot \beta$ i.e. $\mathit{cost}(P'') < \mathit{cost}(P')$. But $\mathit{cost}(P') = \mathit{cost}(P)$ and $\mathit{cost}(P)$ is the optimal cost.  Therefore $\mathit{cost}(P'') < \mathit{cost}(P')$ is contradiction and no such $P''$ exists. \qed
